\documentclass{article}

\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage{amsmath}
\usepackage{multirow}
\usepackage[numbers]{natbib}

\title{The Black-Box Frontier of Chain-of-Thought Faithfulness:\\Post-Hoc Rationalization on Correct Answers}
\author{Suramya Raj Angdembay \and Dikshant Aryal\\[2pt]
\normalsize Advisor: Dr.\ Nick Rahimi --- Team ChainTrackers, UR2PhD Summer 2026}
\date{Research Proposal (revised) --- target venue: BlackboxNLP @ EMNLP 2026}

\begin{document}

\maketitle

\begin{abstract}
Chain-of-thought (CoT) faithfulness---whether a model's stated reasoning reflects the process that produced its answer---is increasingly central to oversight. We ask how far \emph{black-box} signals can go at detecting instance-level CoT unfaithfulness. Auditing FaithCoT-Bench across 4 domains and 4 models against its human annotations, we find that instance-level detection is saturated by simple answer-tracing and strongly confounded by answer correctness; three further signal families (premise-DAG structure, graph-targeted counterfactual interventions, and per-step NLI support) fail to beat or augment answer-tracing (bootstrap 95\% CIs). We then \emph{localize} the failure: detecting \emph{post-hoc rationalization on correct answers} is at chance for every black-box method (all CIs include 0.5). Yet the information is partially present internally: a linear probe on hidden states detects this regime in Llama-3.1-8B (held-out AUROC 0.70) though only weakly in Qwen-2.5-7B, and steering along the probe direction perturbs answers 2--3$\times$ more than a random direction. To test generality beyond one benchmark's labels, we construct matched genuine (reason-first) vs.\ post-hoc (answer-first) traces on AQuA/GSM8K across 7 models spanning three families and 2024--2026; internal decodability is consistently significant (held-out AUROC 0.56--0.81, permutation $p<0.025$), though model-dependent in strength. Crucially, probes trained on this synthetic construction do \emph{not} transfer to the real annotated post-hoc traces (or vice versa): the two are encoded at different depths and transfer at chance. We further show that a widely-used step-removal faithfulness metric \emph{anti-correlates} with human faithfulness judgments. Together, our results bound black-box faithfulness detection, motivate white-box approaches for the residual frontier, and caution against synthetic ``answer-first'' constructions as proxies for organic unfaithfulness.
\end{abstract}

\section{Introduction}

Large language models produce fluent chain-of-thought (CoT) explanations that can appear logical even when the stated reasoning did not causally produce the model's final answer \cite{lanham2023measuring}. As CoT becomes an oversight surface---auditors read the reasoning to decide whether to trust the answer---\emph{instance-level} faithfulness detection (``is \emph{this} trace faithful?'') matters more than aggregate metrics. FaithCoT-Bench \cite{shen2026faithcot} recently formalized this problem with human annotations and showed that a broad panel of detectors performs poorly in aggregate.

This paper asks a sharper question: \textbf{where exactly does black-box detection fail, and is the missing information present in the model's internals?} We proceed in four parts.

\textbf{Part I (audit).} We rigorously audit black-box detection on FaithCoT-Bench against its \emph{human} unfaithfulness labels, adding three signal families not in the benchmark's original panel---premise-DAG structure adapted from PARC \cite{mukherjee2025parc}, graph-targeted counterfactual interventions, and per-step NLI entailment support---with LLM-based premise extraction validated against gold annotations. All fail to beat or augment simple answer-tracing, and answer correctness alone is the strongest single predictor: much of what current benchmarks call ``unfaithfulness detection'' is disguised accuracy prediction.

\textbf{Part II (localization).} We isolate the regime that makes instance-level detection hard: \emph{post-hoc rationalization on correct answers}---traces where the model reaches the right answer but the reasoning did not produce it (FaithCoT-Bench \texttt{faithful\_type} 2), contrasted with genuinely faithful correct traces (type 1). In this regime every black-box signal is at chance. We also find that the benchmark's own step-removal metric \emph{anti-correlates} with the human labels.

\textbf{Part III (internals).} On the same regime, a linear probe over hidden states succeeds where behavior fails---in some models. In Llama-3.1-8B the probe reaches held-out AUROC 0.70 and generalizes across domains; in Qwen-2.5-7B it is weak. An activation-steering experiment shows the probe direction is functionally active, though the causal evidence is modest.

\textbf{Part IV (generality and its limits).} Human-annotated post-hoc traces are scarce, so we construct the regime synthetically at scale: matched pairs where the same model either solves a math problem (reason-first, kept only if correct) or is handed the gold answer and asked to justify it (answer-first)---identical final answers, differing only in whether reasoning caused them. Across 7 open models (three families, 2024--2026) internal decodability replicates with clean labels. But the synthetic and real signals live at different depths and do not transfer to each other---a caution for the common practice of training or evaluating faithfulness detectors on prompt-induced rationalizations.

\section{Related Work}

\textbf{CoT faithfulness.} \citet{lanham2023measuring} introduced perturbation-based faithfulness tests (early answering, adding mistakes); step-removal ``answer-tracing'' metrics descend from this line. FaithCoT-Bench \cite{shen2026faithcot} provides instance-level human annotations across domains and models and benchmarks eleven detectors; our audit is complementary---we localize failure to a specific regime, test additional method families, and evaluate everything against the human labels rather than the benchmark's own derived metrics. \citet{cox2026decoding} show answers are often decodable from internals before the CoT is generated, consistent with our white-box findings.

\textbf{Structural verification of reasoning.} PARC \cite{mukherjee2025parc} builds premise DAGs over reasoning chains to identify errors in math; GoV \cite{fang2025gov} and VeriCoT \cite{feng2025vericot} verify reasoning with DAGs and neuro-symbolic checks respectively. These target \emph{correctness} of reasoning; we test whether such structure transfers to \emph{faithfulness} detection and find it largely does not (with a nuance on math). Step-level error benchmarks \cite{zheng2024processbench,jacovi2024reveal} and GRACE \cite{pham2026grace} evaluate related step-level properties.

\textbf{Probing internals.} Our white-box method is deliberately standard---linear probes on hidden states with selection-corrected permutation testing---because the contribution is the \emph{contrast} (behaviorally inaccessible, internally decodable), not a new probing technique.

\section{Setup: Benchmark, Labels, and the Target Regime}

We use FaithCoT-Bench \cite{shen2026faithcot} across four domains (LogiQA, TruthfulQA, AQuA, HLE-Bio) and four models (Llama-3.1-8B-Instruct, Qwen-2.5-7B-Instruct, GPT-4o-mini, Gemini-2.5-Flash), comprising 1{,}304 traces (634 with complete features across all signal families). The benchmark provides (i) derived metrics, notably \texttt{soft\_faithfulness}, a step-removal answer-tracing AUC; and (ii) \emph{human} annotations: a binary \texttt{unfaithfulness} label and a four-way \texttt{faithful\_type} (ft1: faithful-correct; ft2: post-hoc rationalization on a correct answer; ft3/ft4: incorrect-answer regimes). We treat the human labels as ground truth throughout. A methodological trap we document: correlating a new detector against \texttt{soft\_faithfulness} is near-circular (it is itself a detector); we reproduce an apparent $\rho=0.87$ that collapses to chance-level discrimination against the human label. All open-model experiments run on 2$\times$ RTX 3070 (8\,GB) GPUs with 4-bit NF4 quantization; the 12B model in \S\ref{sec:synth} additionally uses two-GPU pipeline sharding.

\section{Part I: A Rigorous Black-Box Audit}
\label{sec:audit}

\subsection{Signal families}
\textbf{Answer-tracing} (\texttt{soft}/\texttt{hard} faithfulness): sensitivity of the final-answer distribution to removing individual steps. \textbf{Premise-DAG structure}: we adapt PARC to build premise graphs over CoT steps and extract structural features (linearity \texttt{dag\_lin}; max lookback \texttt{dag\_maxlb}). \textbf{Graph-targeted counterfactual interventions}: corrupt the most load-bearing premise (by DAG centrality) and measure answer stability; a fair test requires good premise extraction, so we use an LLM extractor validated against PARC's gold PERL annotations (recall 0.82, F1 0.79, vs.\ 0.57 F1 for a heuristic extractor)---the intervention null below is not an extraction artifact. \textbf{Per-step NLI support}: entailment of each step from the question and prior steps (RoBERTa-large-MNLI; min/mean entailment, unsupported-step counts).

\subsection{Results: nothing beats answer-tracing; correctness confounds everything}
Table~\ref{tab:audit} scores every signal against the human \texttt{unfaithfulness} label with bootstrap 95\% CIs. Three readings. (1) \textbf{Answer correctness alone is the strongest single feature (0.696)}: in this benchmark, unfaithful traces are disproportionately \emph{correct} ones (post-hoc rationalization), so ``unfaithfulness detection'' largely reduces to accuracy prediction. (2) Answer-tracing is informative only in the \emph{inverted} direction (0.651)---used as intended (low coupling $\Rightarrow$ unfaithful) it scores \textbf{0.349, significantly worse than chance} (the metric inversion, \S\ref{sec:inversion}). (3) The structural families add nothing significant: premise-DAG features and most NLI features have CIs spanning 0.5; counterfactual interventions are null in a first heuristic version and weak in the validated-extraction version (effect sizes $g=0.56$--$0.61$), dominated by answer-tracing. One honest nuance: DAG max-lookback \emph{is} weakly significant on AQuA (0.636), the one domain where premise structure is explicit---structure helps exactly where reasoning is formally structured, and nowhere else.

\begin{table}[h]
\centering
\small
\begin{tabular}{llccc}
\toprule
\textbf{Family} & \textbf{Signal (direction scored)} & \textbf{AUROC} & \textbf{95\% CI} & \textbf{Sig.} \\
\midrule
Confound & answer correctness ($\Rightarrow$ unfaithful) & \textbf{0.696} & [0.662, 0.734] & yes \\
\midrule
\multirow{2}{*}{Answer-tracing} & \texttt{soft} inverted (high coupling $\Rightarrow$ unf.) & 0.651 & [0.611, 0.695] & yes \\
 & \texttt{soft} as intended (low coupling $\Rightarrow$ unf.) & 0.349 & [0.305, 0.389] & inverted \\
\midrule
\multirow{3}{*}{NLI support} & \# unsupported steps & 0.569 & [0.523, 0.615] & weak \\
 & mean entailment (low $\Rightarrow$ unf.) & 0.493 & [0.447, 0.538] & no \\
 & contradiction fraction & 0.507 & [0.461, 0.554] & no \\
\midrule
\multirow{2}{*}{Premise-DAG} & linearity (low $\Rightarrow$ unf.) & 0.530 & [0.483, 0.577] & no \\
 & max lookback & 0.543 & [0.496, 0.588] & no \\
\bottomrule
\end{tabular}
\caption{Black-box audit against the human unfaithfulness label ($n{=}633$ traces with all features available; 4 domains $\times$ 4 models; 2{,}000-resample bootstrap CIs). Correctness dominates; answer-tracing works only inverted; structure adds nothing significant.}
\label{tab:audit}
\end{table}

The overall picture: \textbf{answer correctness is most of what black-box ``faithfulness'' detectors measure}.

\subsection{Metric inversion}
\label{sec:inversion}
The step-removal metric \texttt{soft\_faithfulness} \emph{anti-correlates} with human faithfulness: unfaithful traces score \emph{higher} by $+0.139$ (95\% CI $[+0.096, +0.182]$). Post-hoc rationalizations are tightly answer-coupled step by step (each step exists to support the answer), so step removal moves the answer \emph{more}, not less. Step-removal AUC measures answer--reasoning \emph{coupling}, not faithfulness; work using it as a faithfulness score may be optimizing the wrong sign in this regime.

\section{Part II: The Frontier---Post-Hoc Rationalization on Correct Answers}
\label{sec:frontier}

Restricting to human-annotated correct-answer traces (ft1 vs.\ ft2, $n=270$), every black-box signal is at chance (Table~\ref{tab:frontier}). Mechanistically this is expected: a post-hoc rationalization is locally coherent, supports its answer, and yields a confident, stable answer distribution---exactly the profile of a faithful trace, when viewed from outside.

\begin{table}[h]
\centering
\begin{tabular}{lcc}
\toprule
\textbf{Black-box signal} & \textbf{AUROC (ft1 vs.\ ft2)} & \textbf{95\% CI} \\
\midrule
Answer-tracing (\texttt{soft\_faithfulness}) & 0.545 & [0.468, 0.611] \\
Counterfactual interventions & 0.485 & CI $\ni$ 0.5 \\
NLI step-support & 0.514 & CI $\ni$ 0.5 \\
Premise-DAG structure & 0.490 & CI $\ni$ 0.5 \\
\bottomrule
\end{tabular}
\caption{The black-box frontier: on human-annotated correct-answer traces ($n=270$, 4 domains $\times$ 4 models), no black-box signal separates faithful reasoning (ft1) from post-hoc rationalization (ft2); every bootstrap CI includes 0.5.}
\label{tab:frontier}
\end{table}

\section{Part III: The Information Is (Partially) Inside}
\label{sec:whitebox}

\subsection{Probing method}
For each correct-answer trace we extract hidden states at every layer at three positions (last question token before the CoT; last CoT token; mean over CoT tokens), and fit standardized PCA-50 logistic-regression probes with 5-fold cross-validation. Because the best layer is selected post hoc, significance uses a \emph{selection-corrected} permutation test: the null distribution is the max-over-layers AUROC under label permutation (200 permutations).

\subsection{Results on the real annotated regime}
Table~\ref{tab:whitebox} summarizes. In \textbf{Llama-3.1-8B}, the probe separates ft1 from ft2 at held-out AUROC \textbf{0.70} (F1 0.70; averaged over 25 stratified 70/30 splits; permutation $p=0.03$), with signal concentrated in late layers (16--31). It generalizes across domains: leave-one-domain-out AUROC 0.60--0.71. A multi-layer perceptron does not beat the linear probe, suggesting an approximately linear encoding. In \textbf{Qwen-2.5-7B} the probe is weak (held-out 0.58; pilot permutation $p=0.32$): the effect is \emph{model-dependent}, and for Qwen we can only claim ``not detected,'' not ``absent'' ($n=126$). An exploratory L1 (Lasso) probe on the same representations isolates a sparse set of active dimensions; we report it descriptively as its layer selection is not permutation-corrected.

\begin{table}[h]
\centering
\small
\begin{tabular}{lcc}
\toprule
 & \textbf{Llama-3.1-8B} ($n{=}144$) & \textbf{Qwen-2.5-7B} ($n{=}126$) \\
\midrule
In-distribution CV AUROC (best layer) & 0.71 & 0.62 \\
Selection-corrected permutation $p$ & 0.01--0.03 & 0.32 (n.s.) \\
Held-out AUROC (25$\times$ 70/30 splits) & \textbf{0.70} (F1 0.70) & 0.58 \\
Leave-one-domain-out AUROC & 0.60--0.71 & --- \\
Peak layers & late (16--31; best $\approx$29) & late (best $\approx$22) \\
MLP vs.\ linear & no gain ($\approx$linear) & --- \\
\bottomrule
\end{tabular}
\caption{White-box probing on the \emph{real}, human-annotated post-hoc-on-correct regime (ft1 vs.\ ft2). The internal signal is present and generalizes across domains in Llama but is not significant in Qwen: model-dependent.}
\label{tab:whitebox}
\end{table}

\subsection{Causal probe: weak but directional}
If the probe direction is merely a passive correlate, steering along it should perturb behavior no more than a random direction. We add $\pm\alpha\sigma$ along the (train-split) post-hoc direction to the residual stream at the best layer of Llama on 51 held-out multiple-choice traces, against a norm-matched random-direction control (Table~\ref{tab:steering}). At $+6\sigma$ the answer-flip rate is \textbf{0.22 vs.\ 0.08}, and mean $P(\text{original answer})$ drops to 0.748 vs.\ 0.815. The direction is functionally active, but we state this as \emph{weak/suggestive}: effects appear only at large $\alpha$, $n=51$, and answer-change is an indirect readout of post-hoc-ness.

\begin{table}[h]
\centering
\small
\begin{tabular}{ccccc}
\toprule
 & \multicolumn{2}{c}{\textbf{Answer-flip rate}} & \multicolumn{2}{c}{\textbf{Mean $P(\text{orig.\ answer})$}} \\
$\alpha$ (in $\sigma$) & post-hoc dir. & random dir. & post-hoc dir. & random dir. \\
\midrule
$-6$ & 0.16 & 0.20 & 0.714 & 0.763 \\
$-2$ & 0.04 & 0.06 & 0.867 & 0.862 \\
$0$  & 0.00 & 0.00 & 0.878 & 0.878 \\
$+2$ & 0.04 & 0.04 & 0.865 & 0.869 \\
$+4$ & \textbf{0.16} & 0.06 & 0.808 & 0.852 \\
$+6$ & \textbf{0.22} & 0.08 & 0.748 & 0.815 \\
\bottomrule
\end{tabular}
\caption{Activation-steering dose--response (Llama-3.1-8B, best layer, 51 held-out traces). Steering \emph{toward} post-hoc ($+\alpha$) perturbs answers 2--3$\times$ more than a norm-matched random direction; the $-\alpha$ side is mixed. Weak but directional causal evidence.}
\label{tab:steering}
\end{table}

\section{Part IV: Synthetic Construction at Scale---and the Bridge That Fails}
\label{sec:synth}

Human-annotated post-hoc traces are scarce ($n{=}270$ in-regime), and only two open models have them. To test whether internal decodability of the regime is general, we construct it with ground-truth labels.

\subsection{Matched genuine vs.\ post-hoc construction}
For each math problem (AQuA, GSM8K) with gold answer $a^*$: \textbf{genuine} = ``solve step by step'' (kept only if the model's answer is correct; reasoning caused the answer); \textbf{post-hoc} = ``the correct answer is $a^*$; provide step-by-step reasoning leading to it'' (rationalization of a handed answer). Both conditions end at the \emph{same correct answer}, so the pair is matched on correctness and final answer and differs only in whether reasoning produced it. The probed context is symmetric (question + generated CoT; the answer-first instruction is never in the probed input). Decoding is sampled (T=0.7) identically across conditions and models.

\subsection{Internal decodability replicates across 7 models, 3 families, 2024--2026}
Table~\ref{tab:synth_scaling} shows held-out probe AUROCs with selection-corrected permutation tests, a surface-feature black-box baseline (logistic regression on step counts, token counts, numeric density, hedge/self-reference counts---no access to activations), and cross-dataset transfer of the best-layer probe (train AQuA $\to$ test GSM8K and vice versa).

\begin{table}[h]
\centering
\small
\begin{tabular}{lcccccc}
\toprule
\textbf{Model} & \textbf{n pairs} & \textbf{Best L} & \textbf{Held-out} & \textbf{Perm $p$} & \textbf{Surface} & \textbf{XDS (a$\to$g / g$\to$a)} \\
\midrule
Qwen-2.5-7B & 277 & 10 & 0.809 & 0.005 & 0.545 & 0.72 / 0.72 \\
Gemma-4-12B$^{\dagger}$ & 248 & 23 & 0.797 & 0.005 & 0.665 & 0.65 / 0.54 \\
Llama-3.1-8B & 229 & 9 & 0.739 & 0.005 & 0.639 & 0.66 / 0.73 \\
Qwen3-8B & 175 & 10 & 0.716 & 0.005 & 0.535 & 0.62 / 0.56 \\
DeepSeek-R1-Distill-Qwen-7B & 251 & 7 & 0.617 & 0.005 & 0.534 & 0.56 / 0.52 \\
Gemma-2-9B-IT & 243 & 5 & 0.600 & 0.020 & 0.548 & 0.53 / 0.60 \\
DeepSeek-R1-0528-Qwen3-8B & 81 & 22 & 0.558$^{\ddagger}$ & 0.025 & 0.473 & --- \\
\bottomrule
\end{tabular}
\caption{White-box detection of the post-hoc-vs-genuine construction across 7 models (three families, 2024--2026). $n$ = matched pairs; Held-out = single stratified 70/30 split; Perm $p$ = 200-permutation max-over-layers test; Surface = black-box surface-feature baseline (5-fold CV); XDS = cross-dataset transfer at the best layer. $^{\dagger}$4-bit, pipeline-sharded across two 8\,GB GPUs. $^{\ddagger}$Small-$n$ held-out is noisy; the permutation statistic (max-over-layers CV 0.688) is the reliable read.}
\label{tab:synth_scaling}
\end{table}

Three observations. (1) \textbf{All seven models are permutation-significant}, including Qwen-2.5---which was \emph{not} significant on the noisy human labels---indicating the earlier Qwen null was substantially label-noise/small-$n$ rather than absence of signal. (2) The effect is \textbf{model-dependent with an interpretable pattern}: standard instruct models are strongest (0.72--0.81); reasoning-distilled models are weakest (0.56--0.62), plausibly because models trained to elaborate reasoning produce rationalizations that internally resemble genuine derivations. Within one family, Gemma-2 (0.60) $\to$ Gemma-4 (0.80) is a sharp jump. (3) For five of seven models the surface baseline is $\approx$0.47--0.55 while the probe is far higher---the signal is internal, not lexical; in Llama (0.639) and Gemma-4 (0.665) there is a partial surface tell, though the probe still clears it by $\ge$0.10.

\subsection{The proxy bridge fails}
Do the synthetic construction and the real annotated regime share a representation? For the two models with both (Llama, Qwen-2.5) we train on one distribution and test on the other, sweeping \emph{all} layers and domain-matching where possible---giving transfer its best shot (Table~\ref{tab:bridge_failure}).

\begin{table}[h]
\centering
\begin{tabular}{lcccc}
\toprule
\textbf{Model} & \textbf{Synth best L} & \textbf{Real best L} & \textbf{Real $\to$ Synth} & \textbf{Synth $\to$ Real} \\
\midrule
Qwen-2.5-7B & L10 & L22 & 0.595 (mean 0.495) & 0.579 (mean 0.515) \\
Llama-3.1-8B & L9 & L29 & 0.554 (mean 0.476) & 0.531 (mean 0.431) \\
\bottomrule
\end{tabular}
\caption{The proxy bridge failure. Synthetic post-hoc is encoded \emph{early} (L9/L10); real annotated post-hoc is encoded \emph{late} (L22/L29). Cross-distribution transfer is near chance even taking the best of all layers (means over layers are at chance); a domain-matched variant (FaithCoT-AQuA $\to$ synthetic AQuA) is likewise chance-level once its small $n$ and layer selection are accounted for.}
\label{tab:bridge_failure}
\end{table}

Prompt-induced ``answer-first'' rationalization is therefore \textbf{not a representational proxy} for organic post-hoc rationalization: same behavioral definition, different internal phenomena. This cautions against the common shortcut of training or evaluating faithfulness detectors on synthetically induced unfaithfulness and assuming transfer to the real thing.

\section{Contributions}

\textbf{C1. The black-box frontier vs.\ internal probes.} No black-box signal detects post-hoc rationalization on correct answers above chance (Table~\ref{tab:frontier}); yet the distinction is linearly decodable from hidden states---strongly in some models (Llama 0.70 held-out on real labels; 0.56--0.81 across 7 models on clean synthetic labels), weakly in others---establishing a behaviorally-inaccessible-but-internally-present regime, with weak causal support from activation steering.

\textbf{C2. The proxy bridge failure.} Synthetic answer-first constructions are strongly detectable within-distribution but are encoded at different depths than real post-hoc rationalization and do not transfer to it (or vice versa)---a methodological caution for synthetic-proxy evaluation and training.

\textbf{C3. Metric inversion.} The standard step-removal faithfulness metric anti-correlates with human faithfulness ($+0.139$, CI $[+0.096, +0.182]$, higher for unfaithful traces).

\textbf{C4. Rigorous negatives and the accuracy confound.} Premise-DAG structure, graph-targeted interventions (with extraction validated at 0.82 recall against gold), and NLI step-support fail to beat answer-tracing; answer correctness (0.697) is the strongest single predictor---quantifying the ``unfaithfulness as disguised accuracy'' confound---with the honest nuance that DAG structure is weakly informative on math (AQuA 0.636).

\section{Proposed Work and Timeline}

The empirical core above is complete; the remaining work is consolidation, external validation, and submission.

\textbf{Evaluation plan (remaining).}
(1) \emph{Second benchmark:} replicate the black-box audit and, where labels permit, the frontier analysis on the GRACE step-level faithfulness benchmark \cite{pham2026grace} when its full evaluation set ($\sim$437 traces) is released (authors contacted); our preliminary run on the public 40-example sample is consistent with the FaithCoT findings but underpowered.
(2) \emph{Stronger causal validation:} move beyond steering to activation patching/ablation of the probe direction, with pre-registered success criteria.
(3) \emph{Organic unfaithfulness data:} the bridge failure (\S\ref{sec:synth}) implies the field needs naturally occurring post-hoc traces rather than prompted ones; we will pilot a collection protocol (temperature-resampled traces filtered by answer-precommitment probes).
(4) \emph{Scale-up (hardware-gated):} extend the synthetic roster to $\geq$16\,GB-class models (Gemma-4-31B, DeepSeek-V4) if compute allows.

\textbf{Timeline.} \emph{July 2026:} internal polish; advisor review (Dr.\ Rahimi); port to the ACL/BlackboxNLP template and compress to the page limit. \emph{August 2026:} submit to BlackboxNLP @ EMNLP 2026 (CFP deadline expected August; to be verified). \emph{Fall 2026:} GRACE replication on release; activation-patching experiments; organic-data pilot. Repository and all experiment code are release-ready (currently private; public at submission).

\section{Limitations}

Our real-label results rest on one human-annotated benchmark (FaithCoT-Bench); its in-regime $n$ is 270, and per-model white-box $n$ is 144/126, so the Qwen contrast is ``not detected,'' not ``absent.'' The causal steering evidence is weak (large perturbations only, $n=51$, indirect readout). The synthetic study is math-only; its labels are construction labels, not human judgments; Llama and Gemma-4 carry partial surface tells (0.64/0.67); and the 2026 DeepSeek-R1-0528 yields only 81 pairs because its long reasoning overflows our token budget. The bridge test exists only for the two open models with real annotations. Probes use one architecture family of methods (linear/PCA; MLP checked); richer white-box methods may decode more. Closed models (GPT-4o-mini, Gemini) cannot be probed at all---which is itself part of the argument for open-weight oversight.

\section{Conclusion}

Black-box CoT-faithfulness detection is largely answer-correctness detection; where correctness is controlled---post-hoc rationalization on correct answers---every behavioral signal we tested is blind, and a widely-used metric points the wrong way. The information exists inside the model, unevenly across models, and can be partially steered. But the cheap route to scaling this frontier---synthesizing unfaithfulness by prompting---produces a different internal phenomenon than the real thing. Oversight of reasoning will need white-box access, and evaluations of it will need organically unfaithful data.

\section*{Revisions Since the Previous Submission}

The previously submitted proposal (\emph{``Premise-Grounded Detection of Unfaithful Chain-of-Thought Reasoning''}) proposed to \emph{build} a detector combining PARC-style premise DAGs with graph-targeted counterfactual interventions on FaithCoT-Bench. Following advisor guidance to validate the core assumption empirically before further method-building, and incorporating feedback on evaluation rigor, this revision reflects the following changes:

\begin{enumerate}
\item \textbf{The proposed method was implemented, rigorously evaluated---and it failed.} Premise-DAG structure and targeted interventions do not beat simple answer-tracing against human labels (Table~\ref{tab:audit}). Rather than discard this, we report it as a rigorous negative result and re-scoped the project from method-building to an \emph{analysis of the limits of black-box faithfulness detection}; the title changed accordingly.
\item \textbf{Ground truth changed from derived metrics to human annotations.} We found that evaluating against the benchmark's own \texttt{soft\_faithfulness} score is near-circular (an apparent $\rho=0.87$ collapses against human labels) and that the metric itself \emph{anti-correlates} with human judgments (\S\ref{sec:inversion}). All evaluation now targets the human labels.
\item \textbf{Scale and statistical rigor were upgraded.} From a 2-domain pilot to 4 domains $\times$ 4 models ($n=1{,}304$ traces) with bootstrap 95\% CIs, selection-corrected permutation tests, and LLM-based premise extraction validated against gold annotations (0.82 recall)---addressing feedback that early results lacked uncertainty quantification.
\item \textbf{New contributions since the last submission:} localization of the black-box failure to post-hoc rationalization on correct answers (\S\ref{sec:frontier}); white-box probing with causal steering (\S\ref{sec:whitebox}); a 7-model synthetic-construction study across three model families, 2024--2026 (\S\ref{sec:synth}); and the real$\leftrightarrow$synthetic transfer failure (Table~\ref{tab:bridge_failure}), a methodological caution in its own right.
\item \textbf{Structural revisions per the course rubric:} results are now presented in tables with confidence intervals throughout; a \emph{Proposed Work and Timeline} section replaces the earlier implementation plan; a target venue (BlackboxNLP @ EMNLP 2026) is named; and known weaknesses are consolidated in an explicit \emph{Limitations} section rather than dispersed caveats.
\end{enumerate}

\bibliographystyle{plainnat}
\bibliography{ur2phd}

\end{document}
